\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2010-03-10 papers/jss/dpx4Rev.Rnw
Final changes to jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@407 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

	\pkg{depmixS4} implements a general framework for defining and
	estimating dependent mixture models in the \proglang{R}
	programming language \citep{R2009}.  This includes standard Markov
	models, latent/hidden Markov models, and latent class and finite
	mixture distribution models.  The models can be fitted on mixed
	multivariate data with distributions from the \code{glm} family,
	the logistic multinomial, or the multivariate normal distribution.
	Other distributions can be added easily, and an example is
	provided with the exgaus distribution.  Parameters are estimated by
	the EM algorithm or, when (linear) constraints are imposed on the
	parameters, by direct numerical optimization with the
	\pkg{Rdonlp2} or \pkg{Rsolnp} routines.}

@@

\begin{figure}[htbp]
\begin{center}

	<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=
	library(depmixS4)
	data(speed)
	plot(as.ts(speed[1:168,]),main="Speed-accuracy trade-off")
	@

	\caption{Response times (rt), accuracy (corr) and pay-off 
	values (Pacc) for the first series of responses in dataset 
	\code{speed}.}

	\label{fig:speed}

\end{center}
\end{figure}

@@

The EM algorithm however has some drawbacks.  First, it can be slow to
converge towards the end of optimization.  Second, applying
constraints to parameters can be problematic; in particular, EM can
lead to wrong parameter estimates when applying constraints.  Hence,
in \pkg{depmixS4}, EM is used by default in unconstrained models, but
otherwise, direct optimization is used.  Two options are available for
direct optimization using package \pkg{Rdonlp2}
\citep{Tamura2009,Spellucci2002}, or package \pkg{Rsolnp}
\citep{Ghalanos2010}.  Both packages can handle general linear
(in)equality constraints (and optionally also non-linear constraints).

@@

A dependent mixture model is defined by the number of states and the
initial state, state transition, and response distribution functions.
A dependent mixture model can be created with the
\code{depmix}-function as follows:
<<>>=
library(depmixS4)
data(speed)
set.seed(1)
mod <- depmix(response=rt~1, data=speed, nstates=2, 
    trstart=runif(4))
@

@@

\paragraph{Starting values} Note also that start values for the
transition parameters are provided in this call by the
\code{trstart} argument.  At this time, the package does not provide
automatic starting values.  Starting values for parameters can be
provided using three arguments: \code{instart} for the parameters
relating to the prior probabilities, \code{trstart}, relating the
transition models, and \code{respstart} for the parameters of the
response models.  Note that the starting values for the initial and
transition models as well as multinomial logit response models
are interpreted as {\em probabilities}, and internally converted to
multinomial logit parameters (if a logit link function is used).

@@

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The class has \code{print} and
\code{summary} methods to see the results.  The \code{print} method
provides summary information on convergence, the log-likelihood and
the AIC and BIC values:
<<>>=
fm 
@

@@

Since no further arguments were specified, the initial state, state
transition and response distributions were set to their defaults
(multinomial distributions for the first two, and a Gaussian
distribution for the response).

@@

By default, the transition probabilities and the initial state
probabilities are parameterized using a multinomial model with an 
identity link function.
Using a multinomial logistic model allows us to include covariates on the
initial state and transition probabilities. In this case, each row of the 
transition matrix is parameterized by
a baseline category logistic multinomial, meaning that the parameter
for the base category is fixed at zero \citep[see][p.\ 267 ff., for
multinomial logistic models and various
parameterizations]{Agresti2002}.  The default baseline category is the
first state.  Hence, for example, for a 3-state model, the initial
state probability model would have three parameters of which the first
is fixed at zero and the other two are freely estimated. \citet{Chung2007} discuss
a related latent transition model for repeated measurement data
($T=2$) using logistic regression on the transition parameters; they
rely on Bayesian methods of estimation.  Covariates on the transition
probabilities can be specified using a one-sided formula as in the
following example:
<<>>=
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
    transition=~scale(Pacc), instart=runif(2))
fm <- fit(mod,verbose=FALSE)
@

Applying \code{summary} to the \code{fit}ted object gives (only
transition models printed here):
<<>>=
summary(fm, which="transition") 
@

@@

@@

Using package \pkg{Rdonlp2} by \citet{Tamura2009} or \pkg{Rsolnp}
\citet{Ghalanos2010}, parameters may be fitted subject to general linear
(in-)equality constraints.  Constraining and fixing parameters is done
using the \code{conpat} argument to the \code{fit}-function, which
specifies for each parameter in the model whether it's fixed (0) or
free (1 or higher).  Equality constraints can be imposed by giving two
parameters the same number in the \code{conpat} vector.  When
only fixed values are required, the \code{fixed} argument can be used
instead of \code{conpat}, with zeroes for fixed parameters and other
values (ones e.g.) for non-fixed parameters.  Fitting the models
subject to these constraints is handled by the optimization routine
\code{donlp2}.  To be able to construct the \code{conpat} and/or
\code{fixed} vectors one needs the correct ordering of parameters
which is briefly discussed next before proceeding with an example.

After this, we use the fitted values from this model to constrain the
regression coefficients on the transition matrix (parameters number~6
and~10):
<<term=FALSE>>=
# start with fixed and free parameters
conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
conpat[6] <- conpat[10] <- 2
fm2 <- fit(fm1,equal=conpat)
@

Using \code{summary} on the fitted model shows that the regression 
coefficients are now estimated at the same value of 12.66. The function
\code{llratio} computes the likelihood ratio $\chi^2$-statistic and the
associated $p$-value with appropriate degrees of freedom for testing the
tenability of constraints \citep{Dannemann2007}. Note that these arguments 
provide the possibility for arbitrary 
constraints, also between, e.g., a multinomial regression coefficient 
for the transition matrix and the mean of a Gaussian response model. 
Whether such constraints make sense is hence the responsibility of 
the user.

@@

\bibliography{all,ingmar}

>>>>

	\pkg{depmixS4} implements a general framework for defining and
	estimating dependent mixture models in the \proglang{R}
	programming language \citep{R2009}.  This includes standard Markov
	models, latent/hidden Markov models, and latent class and finite
	mixture distribution models.  The models can be fitted on mixed
	multivariate data with distributions from the \code{glm} family,
	the (logistic) multinomial, or the multivariate normal distribution.
	Other distributions can be added easily, and an example is
	provided with the exgaus distribution.  Parameters are estimated by
	the EM algorithm or, when (linear) constraints are imposed on the
	parameters, by direct numerical optimization with the
	\pkg{Rsolnp} or \pkg{Rdonlp2} routines.}

@@

\begin{figure}[htbp]
\begin{center}
<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=	
.Options$digits<-3
library(depmixS4)
data(speed)
plot(as.ts(speed[1:168,]),main="Speed-accuracy trade-off")
@
	\caption{Response times (rt), accuracy (corr) and pay-off 
	values (Pacc) for the first series of responses in dataset 
	\code{speed}.}
	\label{fig:speed}
\end{center}
\end{figure}

@@

The EM algorithm however has some drawbacks.  First, it can be slow to
converge towards the end of optimization.  Second, applying
constraints to parameters can be problematic; in particular, EM can
lead to wrong parameter estimates when applying constraints.  Hence,
in \pkg{depmixS4}, EM is used by default in unconstrained models, but
otherwise, direct optimization is used.  Two options are available for
direct optimization using package \pkg{Rsolnp}
\citep{Ghalanos2010,Ye1987}, or \pkg{Rdonlp2}
\citep{Tamura2009,Spellucci2002}.  Both packages can handle general
linear (in)equality constraints (and optionally also non-linear
constraints).

@@

A dependent mixture model is defined by the number of states and the
initial state, state transition, and response distribution functions.
A dependent mixture model can be created with the
\code{depmix}-function as follows:
<<>>=
library(depmixS4)
data(speed)
set.seed(1)
mod <- depmix(response=rt~1, data=speed, nstates=2, 
    trst=runif(4))
@

@@

\paragraph{Starting values} Note also that start values for the
transition parameters are provided in this call by the \code{trstart}
argument.  Starting values for parameters can be provided using three
arguments: \code{instart} for the parameters relating to the prior
probabilities, \code{trstart}, relating the transition models, and
\code{respstart} for the parameters of the response models.  Note that
the starting values for the initial and transition models as well as
multinomial logit response models are interpreted as {\em
probabilities}, and internally converted to multinomial logit
parameters (if a logit link function is used).  Start values can also
be generated randomly within the EM algorithm by generating random
uniform values for the values of $\gamma_{t}$ in (\ref{eq:Q}) at
iteration 0.  Automatic generation of starting values is not available
for constrained models (see below).

@@

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The class has \code{print} and
\code{summary} methods to see the results.  The \code{print} method
provides information on convergence, the log-likelihood and
the AIC and BIC values:
<<>>=
fm 
@

@@

Since no further arguments were specified, the initial state, state
transition and response distributions were set to their defaults
(multinomial distributions for the first two, and a Gaussian
distribution for the response).  The resulting model indicates two
well-separated states, one with slow and the second with fast
responses.  The transition probabilities indicate rather stable
states, i.e., the probability in either of the states is around 0.9.
The initial state probability estimates indicate that state 1 is the
starting for the process, with a negligible probability of starting in
state 2.

@@

By default, the transition probabilities and the initial state
probabilities are parameterized using a multinomial model with an
identity link function.  Using a multinomial logistic model allows us
to include covariates on the initial state and transition
probabilities.  In this case, each row of the transition matrix is
parameterized by a baseline category logistic multinomial, meaning
that the parameter for the base category is fixed at zero
\citep[see][p.\ 267 ff., for multinomial logistic models and various
parameterizations]{Agresti2002}.  The default baseline category is the
first state.  Hence, for example, for a 3-state model, the initial
state probability model would have three parameters of which the first
is fixed at zero and the other two are freely estimated.
\citet{Chung2007} discuss a related latent transition model for
repeated measurement data ($T=2$) using logistic regression on the
transition parameters; they rely on Bayesian methods of estimation.
Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example: 
<<>>=
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(),
	   transition=~scale(Pacc), instart=runif(2)) 
fm <-fit(mod,verbose=FALSE) 
@

Note the use of \code{verbose=FALSE} to suppress printing of
information on the iterations of the fitting process.  Applying
\code{summary} to the \code{fit}ted object gives (only transition
models printed here by using argument \code{which}): 
<<>>= 
summary(fm, which="transition") 
@

@@

Using package \pkg{Rsolnp} \citet{Ghalanos2010} or \pkg{Rdonlp2} by
\citet{Tamura2009}, parameters may be fitted subject to general linear
(in-)equality constraints.  Constraining and fixing parameters is done
using the \code{conpat} argument to the \code{fit}-function, which
specifies for each parameter in the model whether it's fixed (0) or
free (1 or higher).  Equality constraints can be imposed by giving two
parameters the same number in the \code{conpat} vector.  When only
fixed values are required, the \code{fixed} argument can be used
instead of \code{conpat}, with zeroes for fixed parameters and other
values (ones e.g.) for non-fixed parameters.  Fitting the models
subject to these constraints is handled by the optimization routine
\code{donlp2}.  To be able to construct the \code{conpat} and/or
\code{fixed} vectors one needs the correct ordering of parameters
which is briefly discussed next before proceeding with an example.

After this, we use the fitted values from this model to constrain the
regression coefficients on the transition matrix (parameters number~6
and~10):
<<term=FALSE>>=
pars <- c(unlist(getpars(fm1)))
pars[6] <- pars[10] <- 11
fm1 <- setpars(fm1,pars)
# start with fixed and free parameters
conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
conpat[6] <- conpat[10] <- 2
fm2 <- fit(fm1,equal=conpat)
@

Using \code{summary} on the fitted model shows that the regression
coefficients are now estimated at the same value of 12.667.  The
function \code{llratio} computes the likelihood ratio
$\chi^2$-statistic and the associated $p$-value with appropriate
degrees of freedom for testing the tenability of constraints
\citep{Dannemann2007}.  Note that these arguments (i.e. \code{conpat}
and \code{conrows}) provide the possibility for arbitrary constraints,
also between, e.g., a multinomial regression coefficient for the
transition matrix and the mean of a Gaussian response model.  Whether
such constraints make sense is hence the responsibility of the user.

@@

@@

\bibliography{dpx4Rev}
